Comparison of dimensionality reduction schemes for parallel global optimization algorithms \thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150}

### 2018-04-08 2018/RSD/MultiEvolvents/RSD.tex
RSD: add conclusion

\keywords{Global optimization \and  Dimension reduction \and Parallel algorithms  \and Multidimensional multiextremal optimization  \and Global search algorithms  \and Parallel computations }
\end{abstract}
%
%
%

\section{Introduction}\label{sec:intro}

@@

\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer\cite{globalizerSystem}, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbf{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbf{R}^N\) space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .
\par
As it has been shown earlier by many researchers
(see, for instance, \cite{floudasPardalosGOState}, \cite{horstTuyGO}, \cite{pinterGO}, \cite{strSergGO}),
finding the numerical estimate of the global optimum implies constructing a coverage of
the search domain \(D\). As a result, the computational costs of solving the global
optimization problems are readily very high even for a small number of varied parameters
(the dimensionality of the problem). A notable reduction in the volume of computations
can be achieved when the coverage of the search domain is non-uniform, i. e. the series
of trial points is only dense in a vicinity of the global optimum point. The construction
of such a non-uniform coverage could be provided in an adaptive way, when the selection
of the next trial points is determined by using the search information (the preceding
trial points and the values of the minimized function at these points) obtained in the course
of computations. This necessary condition complicates considerably the computational schemes
of global optimization methods since it implies a complex analysis of a large amount
of multidimensional search information. As a result, many optimization algorithms use
various approaches to the dimensional reduction \cite{pinterGO}, \cite{sergeyevStronginLera2013}, \cite{strongin1978}, \cite{stronginGergelBarkalovParGO}, \cite{strSergGO}.

\section{Methods of Dimension Reduction}
\subsection{Базовый алгоритм}

@@

\subsection{Сдвиговые}
\label{sec:shifted}

@@

\subsection{Вращаемые}
The application of the scheme for building the multiple evolvents (hereinafter called the shifted evolvents or $S$-evolvents) described in Subsection \ref{sec:shifted} allows to preserve the information on the nearness of the points in the multidimensional space and, therefore, to provide more precise (as compared to a single evolvent) estimate of Lipschitz constant in the search process. However, this approach has serious restrictions, which narrow the applicability of the parallel algorithms, designed on the base of the $S$-evolvents.

@@

\subsection{Non injective evolvent}

@@

\subsection{Smooth evolvent}

@@

\subsection{Параллельные множественные отображения}
\label{sec:parallel_evolvents}
Using the multiple mapping allows solving initial problem (1) by parallel solving the problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1≤s≤S
\]
on a set of intervals [0,1] by the index method. Each one-dimensional problem is solved on a separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by particular processor are interpreted as the results of the trials in the rest problems (in the corresponding points \(x^(k_1 ),…,x^(k_S ))\). In this approach, a trial at the point \(x^k \in [0,1]\) executed in the framework of the \(s\)-th problem, consists in the following sequence of operations.
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point\( y^k\) (the blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1≤s≤S\), of the point \(y^k\) and interpret the trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points \(x{}^{k_1} ,…,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par

@@

\section{Results of Numerical Experiments}
The computational experiments have been carried out on the Lobachevsky supercomputer at
State University of Nizhny Novgorod. A computational node included 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. The CPUs had 8 cores (i. e. total 16 cores
were available per a node). Все рассматриваемые алгортимы и развёртки были реализованы на языке C++ в рамках программной системы Globalizer\cite{globalizerSystem}.
Для реализации параллелизма на одном узле использована технология OpenMP, а для параллелизма на несколько узлов --- стандарт MPI.

@@

\subsection{Сравнение последовательных разверток}
С целью понять, обладает ли какой-либо из перечисленных ранее типов развёрток существенным преимуществом над другими, были построены операционные характеристики индексного метода с различными типами развёрток на классах GKLS 2d Simple и GKLS 3d Simple. The global minimum was considered to be found if the algorithm generates a trial point $y^k$ in the $\delta$-vicinity of the global minimizer, i.e. $\left\|y^k-y^\ast\right\|_\infty\leq\delta$. The size of the vicinity was selected as $\delta = 0.01\left\|b-a\right\|_\infty$. In case of GKLS $\delta=0.01$.

@@

\subsection{Параллельные вращаемые развертки}
Чтобы оценить эффективность параллельного алгортима из секции \ref{sec:parallel_evolvents} были проведены вычислительные эксперименты на классах GKLS 4d (Hard, Simple) и GKLS 5d (Hard, Simple). Значения \(r\) во всех экспериментах равно 5.0, размер \(\delta\)-окрестности известного решения увеличен до 0.3. При решении серий задач использовалось до 8 узлов кластера и до 32 вычислительных потоков на каждом узле.

@@

\section{Вывод о целесообразности применения того или иного вида разверток для того или иного вида задач}

>>>>

\keywords{Global optimization \and  Dimension reduction \and Parallel algorithms  \and Multidimensional multiextremal optimization  \and Global search algorithms  \and Parallel computations }
\end{abstract}

%------------------------------------------------------------------------------
\section{Introduction}\label{sec:intro}

@@

%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer\cite{globalizerSystem}, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbf{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbf{R}^N\) space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .
\par
As it has been shown earlier by many researchers
(see, for instance, \cite{floudasPardalosGOState}, \cite{horstTuyGO}, \cite{pinterGO}, \cite{strSergGO}),
finding the numerical estimate of the global optimum implies constructing a coverage of
the search domain \(D\). As a result, the computational costs of solving the global
optimization problems are readily very high even for a small number of varied parameters
(the dimensionality of the problem). A notable reduction in the volume of computations
can be achieved when the coverage of the search domain is non-uniform, i. e. the series
of trial points is only dense in a vicinity of the global optimum point. The construction
of such a non-uniform coverage could be provided in an adaptive way, when the selection
of the next trial points is determined by using the search information (the preceding
trial points and the values of the minimized function at these points) obtained in the course
of computations. This necessary condition complicates considerably the computational schemes
of global optimization methods since it implies a complex analysis of a large amount
of multidimensional search information. As a result, many optimization algorithms use
various approaches to the dimensional reduction \cite{pinterGO}, \cite{sergeyevStronginLera2013}, \cite{strongin1978}, \cite{stronginGergelBarkalovParGO}, \cite{strSergGO}.

%------------------------------------------------------------------------------
\section{Methods of Dimension Reduction}
\subsection{Базовый алгоритм}

@@

%------------------------------------------------------------------------------
\subsection{Сдвиговые}
\label{sec:shifted}

@@

%------------------------------------------------------------------------------
\subsection{Вращаемые}
The application of the scheme for building the multiple evolvents (hereinafter called the shifted evolvents or $S$-evolvents) described in Subsection \ref{sec:shifted} allows to preserve the information on the nearness of the points in the multidimensional space and, therefore, to provide more precise (as compared to a single evolvent) estimate of Lipschitz constant in the search process. However, this approach has serious restrictions, which narrow the applicability of the parallel algorithms, designed on the base of the $S$-evolvents.

@@

%------------------------------------------------------------------------------
\subsection{Non injective evolvent}

@@

%------------------------------------------------------------------------------
\subsection{Smooth evolvent}

@@

%------------------------------------------------------------------------------
\subsection{Параллельные множественные отображения}
\label{sec:parallel_evolvents}
Using the multiple mapping allows solving initial problem (1) by parallel solving the problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1≤s≤S
\]
on a set of intervals [0,1] by the index method. Each one-dimensional problem is solved on a separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by particular processor are interpreted as the results of the trials in the rest problems (in the corresponding points \(x^(k_1 ),…,x^(k_S ))\). In this approach, a trial at the point \(x^k \in [0,1]\) executed in the framework of the \(s\)-th problem, consists in the following sequence of operations.
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point\( y^k\) (the blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1≤s≤S\), of the point \(y^k\) and interpret the trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points \(x{}^{k_1} ,…,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par

@@

%------------------------------------------------------------------------------
\section{Results of Numerical Experiments}
The computational experiments have been carried out on the Lobachevsky supercomputer at
State University of Nizhny Novgorod. A computational node included 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. The CPUs had 8 cores (i. e. total 16 cores
were available per a node). Все рассматриваемые алгортимы и развёртки были реализованы на языке C++ в рамках программной системы Globalizer\cite{globalizerSystem}.
Для реализации параллелизма на одном узле использована технология OpenMP, а для параллелизма на несколько узлов --- стандарт MPI.

@@

%------------------------------------------------------------------------------
\subsection{Сравнение последовательных разверток}
\label{sec:seq_comp}
С целью понять, обладает ли какой-либо из перечисленных ранее типов развёрток существенным преимуществом над другими, были построены операционные характеристики индексного метода с различными типами развёрток на классах GKLS 2d Simple и GKLS 3d Simple. The global minimum was considered to be found if the algorithm generates a trial point $y^k$ in the $\delta$-vicinity of the global minimizer, i.e. $\left\|y^k-y^\ast\right\|_\infty\leq\delta$. The size of the vicinity was selected as $\delta = 0.01\left\|b-a\right\|_\infty$. In case of GKLS $\delta=0.01$.

@@

%------------------------------------------------------------------------------
\subsection{Параллельные вращаемые развертки}
\label{sec:results_parallel}
Чтобы оценить эффективность параллельного алгортима из секции \ref{sec:parallel_evolvents} были проведены вычислительные эксперименты на классах GKLS 4d (Hard, Simple) и GKLS 5d (Hard, Simple). Значения \(r\) во всех экспериментах равно 5.0, размер \(\delta\)-окрестности известного решения увеличен до 0.3. При решении серий задач использовалось до 8 узлов кластера и до 32 вычислительных потоков на каждом узле.

@@

%------------------------------------------------------------------------------
\section{Conclusions}
В работе рассмотрены 5 различных отображений типа кривой Пеано, применяемых для редукции размерности в задачах глобальной оптимизации.
Из предварительного сравнения, проведенного в секции \ref{sec:seq_comp}, можно сделать слудующие выводы:
\begin{itemize}
  \item гладкая и неинъективная развёртки показывают наилучший результат в задачах малой размерности и могут быть успешно применены при решении задач с трудоемкими целевыми функциями. Особенности этих развёрток не позволяют создать на их основе масштабируемый на несколько узлов кластера алгоритм оптимизации.
  \item сдвиговые развёртки привносят большие накладные расходы на работу метода из-за того, что требуется добавления в задачу (\ref{eq:task}) вспомогательного функционального ограничения. Эксперименты показали, что до 95\% поисковой информации приходится на точки, в которых вичисляется только вспомогательное ограничение. Сдвиговые развёртки могут быть использованы в качестве основы для параллельного алгоритма из секции \ref{sec:parallel_evolvents}, однако расходы на обработку дополнительных точек, скорее всего, приведут к малому коэффициенту ускорения от параллельности. Однако, при достаточно трудоёмкой целевой функции их использование может иметь смысл.
  \item вращаемые развёртки обеспечили приемлемую скорость сходимости на задачах малой размерности в последовательном случае. Их использование не приводит к введению дополнительных ограничений, что позволяет построить эффективный параллельный алгоритм на их основе.
\end{itemize}

В секции \ref{sec:results_parallel} приведены результаты численных экспериментов, которые показывают, что алгоритм из раздела \ref{sec:parallel_evolvents} на основе вращаемых развёрток позволяет получить ускорение до 43 раз при решении серии задач с использованием нескольких узлов вычислитеьного кластера. Стоит заметить, что в расматриваемых задачах целевые функции не являются трудоёмкими (среднее время вычисления $10^{-3}$с) и в случае более сложных задач показатели ускорения по времени будут приближаться к ускорению по итерациям.
